Annotation exploration for SCPCP000006, mapping score threshold `r params$mapping_score_thr`

### 2024-10-22 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params: 
  mapping_score_thr: 0.85
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have four columns: barcode, sample_id, compartment, organ
      data.frame(
        compartment = srat$fetal_kidney_predicted.compartment,
        compartment_score = srat$fetal_kidney_predicted.compartment.score,
        organ = srat$fetal_full_predicted.organ, 
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts")
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows() |>
  # Combine with metadata
  dplyr::left_join(metadata, by = c("sample_id" = "scpca_sample_id")) |>
  # Create column for whether the compartment score passes threshold
  dplyr::mutate(pass_mapping_QC = compartment_score > params$mapping_score_thr)
```

@@

@@

Some cells have an excellent mapping score (>0,85) while other performed worse. Let's have a look at the compartments per patient.

@@

p + geom_hline(yintercept = params$mapping_score_thr)

@@

We do not really see a pattern linking the label transfer score and some meta/clinical data. 
For the later CNV analysis, we can apply a threshold of `r params$mapping_score_thr` for the `fetal_kidney_predicted.compartment.score` while selecting for normal cells, i.e. `immune` and/or `endothelial` cells.

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment",
                    split.by = "sample_id")

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(  df = cell_type_df,
                  feature = "ENSG00000081237",
                  group.by = "compartment")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment",
                    split.by = "sample_id")

@@

For that reason, we will first try to run `CopyKAT` on 2-5 samples that have a decent amount (>100 cells when possible) of normal cells (endothelial and immune cells) and for which we are quite confident with the label transfer:

- sample `SCPCS000194` has > `85` % of cells predicted as kidney and `234` + `83` endothelium and immune cells.

- sample `SCPCS000179` has > `94` % of cells predicted as kidney and `25` + `111` endothelium and immune cells.

- sample `SCPCS000184` has > `96` % of cells predicted as kidney and `39` + `70` endothelium and immune cells.

- sample `SCPCS000205` has > `89` % of cells predicted as kidney and `92` + `76` endothelium and immune cells.

- sample `SCPCS0000208` has > `95` % of cells predicted as kidney and `18` + `35` endothelium and immune cells.

>>>>

---
title: "Annotation exploration for SCPCP000006, mapping score threshold `r params$mapping_score_thr`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params: 
  mapping_score_thr: 0.85
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

We also looked into the mapping scores for each of the compartments and explored the potential effects of score thresholds.
In this notebook, we fixed the mapping score threshold to `r params$mapping_score_thr`.

@@

      # Create and return a data frame from the Seurat object with relevant annotations
      # this data frame will have six columns: barcode, compartment, organ, compartment_score, VWF and PTPRC
      data.frame(
        compartment = srat$fetal_kidney_predicted.compartment,
        compartment_score = srat$fetal_kidney_predicted.compartment.score,
        organ = srat$fetal_full_predicted.organ, 
        PTPRC = FetchData(object = srat, vars = "ENSG00000081237", layer = "counts"),
        VWF = FetchData(object = srat, vars = "ENSG00000110799", layer = "counts"),
        umap = srat@reductions$umap@cell.embeddings
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows() |>
  # Combine with metadata
  dplyr::left_join(metadata, by = c("sample_id" = "scpca_sample_id")) |>
  # Create column for whether the compartment score passes threshold
  dplyr::mutate(pass_mapping_QC = compartment_score > params$mapping_score_thr)

# we create a data frames of annotation of cells that pass the mapping score threshold
cell_type_df_pass <- cell_type_df  |>
  dplyr::filter(pass_mapping_QC)

# we create a data frames of annotation of cells that don't pass the mapping score threshold
cell_type_df_nopass <- cell_type_df  |>
  dplyr::filter(!pass_mapping_QC)
```

@@

The vertical line drawn correspods to the threshold explored in the notebook.

@@

Some cells have an excellent mapping score (> `r params$mapping_score_thr`) while other performed worse. Let's have a look at the compartments per patient.

@@

p + geom_hline(yintercept = params$mapping_score_thr)  + coord_cartesian(ylim = c(0.25,1))

@@

We do not really see a pattern linking the label transfer score and some meta/clinical data.

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_pass,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for cells passing the mapping score threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_nopass,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for cells that don't pass the mapping score threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment",
                    split.by = "sample_id")  +
  ggtitle("boxplot of VWF expression for all cells")

@@

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_pass,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for cells passing the mapping score threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_nopass,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for cellsthat are not passing the mapping score threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment",
                    split.by = "sample_id")  +
  ggtitle("boxplot of PTPRC expression for all cells")

### Umap reduction

We look at the umap reduction per sample.

#### All cells

Point are colored per compartment:

- red = endothelial

- green = fetal nephron

- cyan = immune

- purple = stroma

Black cross show cell that do not pass the mapping threshold.

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
  geom_point() +
  geom_point(data = cell_type_df_nopass, mapping = aes( x = umap.umap_1, y = umap.umap_2),  shape = 4, color = "black", alpha = 0.3) +
  facet_wrap(facets = ~sample_id)
```

#### Cells passing the mapping score threshold

Point are colored per compartment. 
Here we only plotted cells that do pass the mapping score threshold.

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df_pass, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
  geom_point() +
  facet_wrap(facets = ~sample_id)

@@

For that reason, we will first try to run `CopyKAT` on 2-5 samples that have a majority of cells mapping to kidney and a decent amount (>100 cells when possible) of normal cells (endothelial and immune cells) and for which we are quite confident with the label transfer:

- sample `SCPCS000194`

- sample `SCPCS000179`

- sample `SCPCS000184`

- sample `SCPCS000205`

- sample `SCPCS0000208`
